---
title: "Calculation Functions"
author: "FlowWest"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculation Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=680/72,
  fig.height=680*(2/3)/72,
  dpi=72
)
ggplot2::theme_set(ggplot2::theme_minimal())
library(tidyverse)
library(sf)
library(fishFoodMWD)
```

We begin with the `fields` dataset, which contains the rice field geometries and their areas. 

```{r echo=FALSE}
fields |> head(10) |> knitr::kable()
```

Pass the fields to the `calc_inv_mass` function along with the desired number of days.

```
fields |> calc_inv_mass(14)
```
```{r echo=FALSE}
fields |> calc_inv_mass(14) |> head(10) |> knitr::kable()
```

Shorthand to plot the result

```{r}
plot_inv_mass(14)
```

Moss et al. (2009) estimated that invertebrate biomass in flooded post-harvest agricultural fields increased at an average rate of  0.186 g/m^2 per day. 

How much invertebrate biomass is produced by each field in 14 days?

```{r calc-mass-by-field, echo=TRUE, message=FALSE}
fields |> 
  calc_inv_mass(14) |>
  ggplot() + 
    geom_sf(aes(fill = total_prod_kg, color = total_prod_kg)) + 
    geom_sf(data=streams) + geom_sf(data=canals) + geom_sf(data=returns) +
    scale_fill_viridis_c(aesthetics = c("colour", "fill"),
                                  option="cividis",
                                  direction=-1,
                                  name="Total 14-day production (kg) \nby field") +
    ggplot2::scale_y_continuous(breaks = seq(38, 40, by=0.5)) +
    ggplot2::scale_x_continuous(breaks = seq(-122.5, -120.5, by=0.5))
```

Which watersheds produce the most invertebrate biomass?

```{r calc-mass-by-watershed, echo=TRUE, message=FALSE}
total_prod_by_group <- fields |> 
  calc_inv_mass(14) |>
  group_by(group_id) |> 
  summarize(sum_total_prod_kg = sum(total_prod_kg)) |>
  st_drop_geometry()

watersheds |> 
  left_join(total_prod_by_group) |>
  ggplot() + 
    geom_sf(aes(fill = sum_total_prod_kg, color = sum_total_prod_kg)) + 
    geom_sf(data=streams) + geom_sf(data=canals) + geom_sf(data=returns) +
    scale_fill_viridis_c(aesthetics = c("colour", "fill"),
                                  option="cividis",
                                  direction=-1,
                                  name="Total 14-day production (kg) \nby watershed") +
    ggplot2::scale_y_continuous(breaks = seq(38, 40, by=0.5)) +
    ggplot2::scale_x_continuous(breaks = seq(-122.5, -120.5, by=0.5))
```

These results are directly proportional to field acreage.

Growth in biomass over time by field (illustrative example):

```{r calc-field-ts, echo=TRUE, message=FALSE}
fields |> 
  head(n = 10) |>
  calc_inv_mass_ts(14) |>
  ggplot() + geom_line(aes(x=day, y=total_prod_kg, color=unique_id))
```
