---
title: "Modeling Approach and Methods"
author: "FlowWest"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling Approach and Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=680/72,
  fig.height=680*(2/3)/72,
  dpi=72
)
ggplot2::theme_set(ggplot2::theme_minimal())
library(tidyverse)
library(sf)
library(fishFoodMWD)
```

To document and illustrate here:

# Types of flow

Separating the watersheds into groups (show plot of watersheds with return types classified)

```{r}
watersheds_returns <- watersheds |> 
  left_join(st_drop_geometry(returns), by=join_by(return_id))

ggplot() + 
  geom_sf(data=watersheds_returns, aes(fill=return_direct), color="white") +
  geom_sf(data=streams, aes(color="Direct")) + 
  geom_sf(data=canals, aes(color="Indirect")) + 
  geom_sf(data=returns, aes(color=return_direct)) + 
  scale_color_manual(values=c("Direct"="mediumblue", "Indirect"="darkred")) + 
  scale_fill_manual(values=c("Direct"="lightblue", "Indirect"="pink"), na.value="moccasin")
```

* Direct returns:

```{r}
ggplot() + 
  geom_sf(data=filter(watersheds_returns, return_direct=="Direct"), fill="lightblue", color="white") +
  geom_sf(data=filter(returns, return_direct=="Direct"), color="mediumblue") +
  geom_sf(data=streams, color="mediumblue")
```

* Indirect:

```{r}
ggplot() + 
  geom_sf(data=filter(watersheds_returns, return_direct=="Indirect"), fill="pink", color="white") +
  geom_sf(data=filter(returns, return_direct=="Indirect"), color="darkred") +
  geom_sf(data=canals, color="darkred") +
  geom_sf(data=streams, color="mediumblue")
```

Lateral:

```{r}
ggplot() + 
  geom_sf(data=filter(watersheds_returns, is.na(return_direct)), fill="moccasin", color="white") +
  geom_sf(data=streams, color="mediumblue")
```

# Distance calculations

Selected watersheds are calculated with a direct flow to the nearest FBS rather than via a return point

```{r}
fields |> 
  left_join(distances, by=join_by(unique_id)) |>
  filter(!(return_id>0)) |>
  ggplot() + 
    geom_sf(aes(fill=return_dis/5280), color=NA) + 
    scale_fill_gradient(low="moccasin", high="tomato") + 
    geom_sf(data=streams)
```


Distance from fields to return points are calculated for all returns 
```{r}
fields |> 
  left_join(distances, by=join_by(unique_id)) |>
  filter(return_id>0) |>
  ggplot() + 
    geom_sf(aes(fill=return_dis/5280), color=NA) + 
    scale_fill_gradient(low="moccasin", high="tomato") + 
    geom_sf(data=returns) +
    geom_sf(data=canals) +
    geom_sf(data=streams)
```

Flow distance from return to nearest FBS is calculated for the returns that don't directly flow to a FBS
```{r}
watersheds_returns |> 
  filter(return_direct=="Indirect") |>
  group_by(return_id, ds_return_id, ds_fbs_dist) |> 
  summarize() |> 
  ggplot() + 
    geom_sf(aes(fill=ds_fbs_dist/5280), color="white") +
    scale_fill_gradient(low="moccasin", high="tomato") + 
    geom_sf(data=filter(returns, return_direct=="Indirect"), color="darkred") +
    geom_sf(data=canals, color="darkred") +
    geom_sf(data=streams, color="mediumblue")
```

Results are all added together

```{r}
fields |> 
  left_join(distances, by=join_by(unique_id)) |>
  ggplot() + 
    geom_sf(aes(fill=totdist_ft/5280), color=NA) + 
    scale_fill_gradient(low="moccasin", high="tomato") + 
    geom_sf(data=returns) +
    geom_sf(data=canals) +
    geom_sf(data=streams)
```

# Other 

Also document the two different distance methods : straight-line and rectangular
